Comparing Machine Learning and Deep Learning Models with Attention Mechanisms for Multiple Source Code Authorship Attribution\\ \thanks{Identify applicable funding agency here. If none, delete this.}

### 2025-03-14 my_cst/article.tex
m

This study systematically compares machine learning and deep learning 
models, including Random Forests, BERT trained on tokenized source code, 
BERT trained on AST representations, and an Attention-based Neural Network 
(AttentionNN) utilizing AST-derived node-to-node paths. The evaluation is 
conducted on a dataset of 3,004 C language functions from Google Code Jam 
(GCJ) spanning 2009 to 2018 \cite{gcj}. Most of the works in the field
predict the authors based on files consisting of multiple language constructs, our 
research however focuses on prediction of author given single function.
The models are assessed based on their accuracy, precision, recall, and training efficiency across different 
author set sizes.

@@

As we have mentioned, the models were mainly tested for their scalability with having more functions per author while decreasing the distinct author set size.
Table \ref{tab:comparison_models} summarizes how each model performed in terms of accuracy and the time required for training (until the validation loss stopped decreasing) across diverse datasets
and figure \ref{fig:3d_bars} visualizes the achieved performances on Accuracy, Precision and Recall. From table \ref{tab:comparison_models} we can say that BERT 
trained on tokens extracted from the raw source code performed the best at each of the author set sizes, especially with accuracy over 90\% on the smallest set, however with the longest required time to train the model, more than 
9.5 times longer than it took to train AttentionNN to its full potential and 36.4 times longer than the time required for Random Forests which scored just 4.8 percent less and reached similar to state-of-the art results \cite{rf1} \cite{rf2}
What also interested us is that From \texttt{set\_size=11} to \texttt{set\_size=3}, BERT improved only by 0.86\% which is the smallest improvement across all the experiments and indicates that increasing the minimum function 
count per author from 50 to 58 had little impact on the model's ability to distinguish authors. 
On the other hand, from being the worst, BERT fed with AST pre-order traversal scored a 40\% higher under these conditions. Unfortunately, we were unable to find any relevant sources against which to compare our BERT results.
Nevertheless the following experiments conducted on other models, helped us understand advantages and disadvantages of BERT model for authorship attribution.

@@

From the figure \ref{fig:misclass_lens} the only statement we can say is that the model AttentionNN particularly struggled with functions of length lower than 40 tokens
but did quite well on the longer ones. This will be discussed in the following subsection more deeply \ref{subsec:comp_over_quant}.
Other models performance is not really distinguishable from this visualization, so we performed the same index strategy with the Halstead complexity metrics.
Measuring each misclassified function's Halstead complexity and plotting it for Random Forests and BERT trained on tokenized source code, we obtained following figure.

@@

@@

This way we obtain visual representation that objectively accounts for the fact that the number of errors is different.
In the following figure we visualized the frequencies of misclassified function lengths in tokens for Random Forests, BERT and AttentionNN scaling each 
model with its index.

From the figure we can deduce, there is an emerging trend in accuracy from bins of lower value to the bins of higher value.
Eventhough the frequency distribution of both variables indicates more frequency on the lower values we observe a lower
accuracy. This tells that the model is learning rather from more structural complex functions than simple non-nesting and 
control-flow divergent ones.

\begin{figure}[!b]
    \centering
    \hspace*{-25pt}\includegraphics[width=1.2\columnwidth]{figures/misclass_lens.pdf}
    \caption{Frequency distribution of misclassified function lengths (in tokens) for Random Forests, BERT, and AttentionNN.}
    \label{fig:misclass_lens}
\end{figure}

\begin{figure}[!b]
    \centering
    \hspace*{-25pt}\includegraphics[width=1.2\columnwidth]{figures/misclass_hs.png}
    \caption{Frequency distribution of misclassified function Halstead complexity for Random Forests and BERT.}
    \label{fig:misclass_halstead}
\end{figure}

@@

Going forward, it would be interesting to combine these methods. For example a hybrid approach using BERT embeddings with Random Forest 
classifier could balance accuracy and efficiency. Also, using graph-based neural networks to process ASTs more effectively might push 
authorship attribution even further. Exploring data augmentation techniques or better regularization strategies could also help deep 
learning models generalize better, especially with AST-based inputs.

\begin{figure}[!b]
    \centering
    \begin{subfigure}{\columnwidth}
        \centering
        \includegraphics[width=\linewidth]{figures/attention_nn/ann_acc_num_nodes.pdf}
        \caption{Accuracy vs. number of AST nodes}
        \label{fig:ann_acc_nodes}
    \end{subfigure}
    \hfill
    \begin{subfigure}{\columnwidth}
        \centering
        \includegraphics[width=\linewidth]{figures/attention_nn/ann_acc_depth.pdf}
        \caption{Accuracy vs. AST depth}
        \label{fig:ann_acc_depth}
    \end{subfigure}
    \caption{AttentionNN accuracy for functions with different structural complexities.}
    \label{fig:ann_acc}
\end{figure}

@@

>>>>

This study systematically compares machine learning and deep learning 
models, including Random Forests, BERT trained on tokenized source code, 
BERT trained on AST representations, and an Attention-based Neural Network 
(AttentionNN) utilizing AST-derived node-to-node paths. The evaluation is 
conducted on a dataset of 3,756 C language functions from Google Code Jam 
(GCJ) spanning 2009 to 2018 \cite{gcj}. Most of the works in the field
predict the authors based on files consisting of multiple language constructs, our 
research however focuses on prediction of author given single function.
The models are assessed based on their accuracy, precision, recall, and training efficiency across different 
author set sizes.

@@

As we have mentioned, the models were mainly tested for their scalability with having more functions per author while decreasing the distinct author set size.
Table \ref{tab:comparison_models} summarizes how each model performed in terms of accuracy and the time required for training (until the validation loss stopped decreasing) across diverse datasets
and figure \ref{fig:3d_bars} visualizes the achieved performances on Accuracy, Precision and Recall. From table \ref{tab:comparison_models} we can say that BERT 
trained on tokens extracted from the raw source code performed the best at each of the author set sizes, especially with accuracy over 90\% on the smallest set, however with the longest required time to train the model, more than 
9.5 times longer than it took to train AttentionNN to its full potential and 36.4 times longer than the time required for Random Forests which scored just 4.8 percent less. Our RF experiment on dataset with 110 reached results similar to this study \cite{rf1}, 
where they had 525 authors, but with 2-11 files per author and also included information from comments which we filtered out. In \cite{rf2}, authors reached accuracy of 84.5\% on 50 authors with and average of 42.8 files each similar to our setup for 
27 authors. They scored 14\% more than we did, using the the features presented in \cite{caliskan}.

What also interested us is that From \texttt{set\_size=11} to \texttt{set\_size=3}, BERT improved only by 0.86\% which is the smallest improvement across all the experiments and indicates that increasing the minimum function 
count per author from 50 to 58 had little impact on the model's ability to distinguish authors. 
On the other hand, from being the worst, BERT fed with AST pre-order traversal scored a 40\% higher under these conditions. Unfortunately, we were unable to find any relevant sources against which to compare our BERT results.
Nevertheless the following experiments conducted on other models, helped us understand advantages and disadvantages of BERT model for authorship attribution.

@@

This way we obtain visual representation that objectively accounts for the fact that the number of errors is different.
Figure \ref{fig:misclass_lens} shows visualized frequencies of misclassified function lengths in tokens for Random Forests, BERT and AttentionNN scaling each 
model with its index.

From the figure \ref{fig:misclass_lens} the only statement we can say is that the model AttentionNN particularly struggled with functions of length lower than 40 tokens
but did quite well on the longer ones. This will be discussed in the following subsection more deeply \ref{subsec:comp_over_quant}.
Other models performance is not really distinguishable from this visualization, so we performed the same index strategy with the Halstead complexity metrics in figure \ref{fig:misclass_halstead}.
Measuring each misclassified function's Halstead complexity and plotting it for Random Forests and BERT trained on tokenized source code, we obtained following figure.

@@

This model showed relatively promising results, indicating increasing dataset size and providing more complex data enhances its performance.
Paper \cite{pbnn} presents similar model reaching 0.979\% accuracy with 40 authors but 3,021 files for each in programming language Java.

@@

From the figure \ref{fig:ann_acc} we can deduce, there is an emerging trend in accuracy from bins of lower value to the bins of higher value.
Eventhough the frequency distribution of both variables indicates more frequency on the lower values we observe a lower
accuracy. This tells that the model is learning more about the author from structural complex functions than simple non-nesting and 
control-flow divergent ones.

\begin{figure}[!t]
    \centering
    \hspace*{-25pt}\includegraphics[width=1.2\columnwidth]{figures/misclass_lens.pdf}
    \caption{Frequency distribution of misclassified function lengths (in tokens) for Random Forests, BERT, and AttentionNN.}
    \label{fig:misclass_lens}
\end{figure}

\begin{figure}[!t]
    \centering
    \hspace*{-25pt}\includegraphics[width=1.2\columnwidth]{figures/misclass_hs.png}
    \caption{Frequency distribution of misclassified function Halstead complexity for Random Forests and BERT.}
    \label{fig:misclass_halstead}
\end{figure}

@@

Going forward, it would be interesting to combine these methods. For example a hybrid approach using BERT embeddings with Random Forest 
classifier could balance accuracy and efficiency. Also, using graph-based neural networks to process ASTs more effectively might push 
authorship attribution even further \cite{dependence_graphs}. Exploring data augmentation techniques or better regularization strategies could also help deep 
learning models generalize better, especially with AST-based inputs.

\begin{figure}[!t]
    \centering
    \begin{subfigure}{\columnwidth}
        \centering
        \includegraphics[width=\linewidth]{figures/attention_nn/ann_acc_num_nodes.pdf}
        \caption{Accuracy vs. number of AST nodes}
        \label{fig:ann_acc_nodes}
    \end{subfigure}
    \hfill
    \begin{subfigure}{\columnwidth}
        \centering
        \includegraphics[width=\linewidth]{figures/attention_nn/ann_acc_depth.pdf}
        \caption{Accuracy vs. AST depth}
        \label{fig:ann_acc_depth}
    \end{subfigure}
    \caption{AttentionNN accuracy for functions with different structural complexities.}
    \label{fig:ann_acc}
\end{figure}

@@

    \bibitem{caliskan}
    A. Caliskan-Islam \textit{et al.}, ``De-anonymizing programmers via code stylometry,'' in \textit{Proc. 24th USENIX Security Symp.}, 2015, pp. 255--270. [Online]. Available: https://www.usenix.org/conference/usenixsecurity15/technical-sessions/presentation/caliskan-islam

    \bibitem{dependence_graphs}
    R. Chen, L. Hong, C. Chunyan Lü and W. Deng, ``Author Identification of Software Source Code with Program Dependence Graphs,'' in \textit{Proc. 2010 IEEE 34th Annual Computer Software and Applications Conference Workshops}, Seoul, Korea (South), 2010, pp. 281--286, doi: 10.1109/COMPSACW.2010.56. [Online]. Available: https://doi.org/10.1109/COMPSACW.2010.56
